---
title: "Sampling Paths from Gaussian stochastic process"
author: "XC"
date: "2022-06-09"
output: html_notebook
---

```{r load packages, include=F}
install.packages("MASS")
library(MASS)

install.packages('tinytex')
library(tinytex)
```


## General set-up

* a zero-centered GP(m(.), K(.,.)) 
* f is a function drawn from this GP
* a vector (x1, ... xn), the function values (f(x1), ... f(xn)) or (Y)
* must have a multivariate GP with
  - mean: (m(x1),..., m(xn))
  - covariance matrix Sigma with Sigma_{ij}  = K(xi, xj)

* so we could make use of this property (with 2 moments determine the whole distribution), draw this function from GP:
  1. select a fine grid of x-coords
  2. use mvrnorm() from MASS to draw function values at these points
  3. then connect them with straight lines
  
  
  
## Generate Covariance matrix

Generate Covariance matrix from a known kernel function at points x

```{r}
cov_matrix <- function(x, kernel_fn, ...) {
  outer(x, x, function(a, b) kernel_fn(a, b, ...))
}

```


## Sample

Given x-coords, take N draws from the GP with 
covariance K(.,.) evaluated using kernel_fn at coords x

```{r}
draw_samples <- function(x, N, kernel_fn, ...) {
  set.seed(2022-06-09)
  
  K <- cov_matrix(x, kernel_fn, ...)
  
  Y <- matrix(NA, nrow = length(x), ncol = N)
  for (n in 1:N) {
    Y[, n]<- mvrnorm(1, mu = rep(0, length(x)), Sigma = K)
  }
  Y
}

```


## Parameters

Use the following parameters for the rest code
```{r}
x <- seq(0, 2, length.out = 201)  # 1-d simulation

N <- 3

col_list <- list("red", "blue", "black")
```


## Kernel 1:
## Squared exponential(SE) kernel

* SE also known as _radial basis furntion (RBF) kernel_ or the _Gaussian kernel_ 
* has the form:
$K(x, x') = \sigma^2 exp(- \frac{||x-x'||^2}{2l^2} )$
$K(x, x') = \sigma^2 exp(- \frac{\| x-x'\|^2}{2l^2}) $

* where $\sigma^2 > 0$ and $l > 0$ are hyperparameters:
  - $\sigma^2 > 0$ tells how variable the function is overall, and set to 1 for simplicity
  - $l = 0.2$ 
  
* most commonly used kernel as its computational tractability

Now generates 3 draws from the SE kernel
```{r}
## generate SE kernel
SE_kernel <- function(x, y, Sigma = 1, l = 1) {
  Sigma^2 * exp(- (x - y)^2 / (2 * l^2))
}
```

```{r}
## draw samples from GP that use SE_kernel to construct cov_matrix at coords x
Y <- draw_samples(x = x, N = N, kernel_fn = SE_kernel, l = 0.2)

```

```{r}
## plot the sample path

plot(range(x), range(Y), xlab = "x", ylab = "y",
     type = "n", main = "SE_kernel, l = 0.2")

for (n in 1:N) {
  lines(x, Y[, n], col = col_list[[n]], lwd = 1.5)
}

```


## Show how changing "length-scale" parmaeter $l$ affects the function drawn

The smaller the $l$, the more wiggly the function drawn

```{r}
par(mfrow = c(1, 3))

for (l in c(0.2, 0.9, 1.5)) {
  # draw samples from kernel with different l
  Y <- draw_samples(x, N, kernel_fn = SE_kernel, l = l)
  
  # plot 3 frameworks
  plot(range(x), range(Y), xlab = "x", ylab = "y", 
       main = paste0("SE_kernel,", "l = ", l, sep = " "))
       
  # plot sample path
  for (n in 1:N) {
    lines(x, Y[, n], col = col_list[[n]], lwd = 1.5)
  }
  
}
```