---
title: "Sampling Paths from a GP"
author: "XC"
date: "03/07/2021"
output: pdf_document
---


```{r load packages, include=F}
#install.packages("MASS")
#library(MASS)

#install.packages('tinytex')
#library(tinytex)


```

```{r}
#tinytex::install_tinytex()
```

## General set-up

* a zero-centered GP(m(.), K(.,.)) 
* f is a function drawn from this GP
* a vector (x1, ... xn), the function values (f(x1), ... f(xn)) or (Y)
* must have a multivariate GP with
  * mean: (m(x1),..., m(xn))
  * covariance matrix Sigma with Sigma_{ij}  = K(xi, xj)
  
* so we could make use of this property (with 2 moments determin the whole distribution), draw this function from GP:
  1. select a fine grid of x-coords
  2. use mvrnorm() from MASS to draw function values at these points
  3. then connect them with straight lines
  
  
## Generate Covariance matrix

Generate Covariance matrix from a known kernel function at points x
```{r}
cov_matrix <- function(x, kernel_fn, ...) {
  outer(x, x, function(a, b) kernel_fn(a, b, ...))
}
```


## Sample

Given x coords, take N draws from the GP with K evaluated using kernel_fn at x
```{r}
draw_samples <- function(x, N, kernel_fn, ...) {
  set.seed(03-07-2021)
  
  Y <- matrix(NA, nrow = length(x), ncol = N)
  for(n in 1:N) {
    K <- cov_matrix(x, kernel_fn, ...)
    Y[, n]<- mvrnorm(1, mu = rep(0, length(x)), Sigma = K)
  }
  Y
}
```


## Parameters

Use the following parameters for the rest code
```{r}
x <- seq(0, 2, length.out = 201)           # x-coords
N <- 3                                     # no. of draws
col_list <- list("red", "blue", "black")   # col for lines of different draws
```



## Squared exponential(SE) kernel

SE also known as _radial basis furntion (RBF) kernel_ or the _Gaussian kernel_ has the form:
$K(x, x') = \sigma^2 exp(- \frac{||x-x'||^2}{2l^2} )$,

where $\sigma^2 > 0$ and $l > 0$ are hyperparameters. 
$\sigma^2 > 0$ tells how variable the function is overall, and set to 1 for simplicity.

It's the most commonly used kernel as its computational tractability. 

Now generates 3 draws from the SE kernel with $l=0.2$
```{r}
se_kernel <- function(x, y, Sigma = 1, Length = 1) {
  Sigma^2 * exp(- (x - y)^2 / (2 * Length^2))
}


Y <- draw_samples(x, N, kernel_fn = se_kernel, Length = 0.2)

plot(range(x), range(Y), xlab = "x", ylab = "y", type = 'n', 
     main = "se_kernel, Length l = 0.2")


for(n in 1:N) {
  lines(x, Y[, n], col = col_list[[n]], lwd = 1.5)
}

```


## Show how changing "length-scale" parmaeter l affects the function drawn. 

The smaller the $l$, the more wiggly the function drawn

```{r}
par(mfrow = c(1, 3))

for(l in c(0.2, 0.7, 1.5)) {
  Y <- draw_samples(x, N, kernel_fn = se_kernel, Length = l)  # draw samples
  
  plot(range(x), range(Y), xlab = "x", ylab = "y", type = "n",
       main = paste("SE kernel, Length=", l))                 # plot framework
  
  for(n in 1:N) {
    lines(x, Y[, n], col = col_list[[n]], lwd = 1.5)          # lines up samples
  }
  
}


```


```{r}

for(l in c(0.2, 0.7, 1.5)) {
  Y <- draw_samples(x, N, kernel_fn = se_kernel, Length = l)  # draw samples
  str(Y)
}
```
